Tracing the Hercules stream around the Galaxy

### 2010-05-26 tex/hercules.tex
groups-simulations:most of Hogg's comments

\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

\begin{abstract}
It has been proposed that the Hercules stream, a group of co-moving
stars in the Solar neighborhood offset from the bulk of the velocity
distribution, is the result of resonant interactions between stars in
the outer disk and the Galactic bar. So far it has only been seen in
the immediate Solar neighborhood. I predict the distribution of
stellar velocities and the run of the Hercules stream as a function of
location in the Galactic disk in a simple model for the Galaxy and the
bar that produces the observed Hercules stream. I further show that
the Hercules feature is strong enough to be unambiguously detected in
the distribution of line-of-sight velocities in selected directions. I
quantify the most promising lines-of-sight for detection in radial
velocities; these directions are all at $l \gtrsim 250^{\circ}$. The
predictions presented here are only weakly affected by distance
uncertainties, assumptions about the distribution function of the
stellar disk, and the details of the Galactic potential including the
effect of spiral structure. \Gaia\ and future spectroscopic surveys of
the Galactic disk will be able to robustly test the origin of the
Hercules stream and constrain the properties of the Galactic bar.
\end{abstract}

@@

The velocity distribution of nearby stars contains various large
overdensities of co-moving stars that cannot be explained as the
vestiges of the clumpy nature of star formation
\citep[\eg,][]{Bovy10a}. In particular, the Hercules moving group, a
group of stars significantly offset from the bulk of the observed
velocity distribution \citep[\eg,][see
  \figurename~\ref{fig:obs}]{Dehnen98b,Bovy10a}, displays a wide range
of ages and metallicities
\citep{Blaauw70a,raboud98a,caloi99a,Bensby07a}. A compelling model
that has the Hercules stream originating through resonant interactions
of stars in the outer disk with the bar in the central region of the
Galaxy has been put forward \citep{dehnen00a,fux01a}, but any
confirmation of this picture beyond the locally observed stellar
kinematics remains lacking. An investigation of the Hercules members'
color--magnitude diagram found no evidence for any significant
metallicity anomaly with respect to other local stars \citep{Bovy10a},
even though such a difference should be expected as the Hercules stars
originate from a few kiloparsec toward the Galactic center where the
average metallicity is higher than in the Solar neighborhood. The
absence of the predicted metallicity anomaly could potentially be
explained, \eg, through radial mixing \citep{sellwood02a}, but it is
clear that chemodynamical modeling of the Galactic disk is currently
not up to the task of separating out these effects.

In this Letter I propose that a robust and unambiguous test of the
bar-resonance model for the Hercules stream is the distinct pattern it
predicts as we trace the velocity distribution of stars near the Solar
radius around the Galaxy. \Gaia\ is in a good position to have the
final word on this, but I also show that the Hercules feature can be
detected in the line-of-sight velocity distribution in selected
directions on the sky. Clear signatures should exist in regions a few
kiloparsec from the Sun, allowing for a clean verification of the
bar-resonance model for the Hercules stream.

@@

To simulate the effect of the bar on stellar orbits near the Solar
circle, we follow the approach of \citet{dehnen00a}. For a fixed
position in the Galaxy, we evaluate the velocity distribution function
at values for the radial---toward the Galactic center---and
tangential---in the direction of Galactic rotation---velocity by
backward-integrating these velocities in the Galactic potential to
obtain the intial orbit before bar formation. We posit that at this
epoch the old stellar disk was in a steady-state that can be described
by a simple distribution function that is a function of the energy and
angular momentum of the orbit alone. It follows from the collisionless
Boltzmann equation that the value of the velocity distribution at the
present position and velocity is equal to that of the steady-state
distribution function before bar formation evaluated at the initial
orbit.

Since the Hercules stream is only apparent in the planar motions in
the disk \citep[\eg,][]{Bovy09a}, we only consider a two-dimensional
model for the Galaxy. We use a simple power-law rotation curve to
model the axisymmetric Galactic potential
\begin{equation}
v_c(R) = \vo(R/\Ro)^\beta \,,
\end{equation}
and for most of the simulations below we specify this model further to
a flat rotation curve ($\beta = 0$). The model for the bar is given by
the following potential
\begin{equation}
\Phi_b(R,\phi) = \Ab(t) \cos [2(\phi - \Omegab t)] \times
\left\{ \begin{array}{ll} -(\Rb/R)^3\,, & \mathrm{for}\ R \geq
  \Rb\,,\\ (R/\Rb)^3-2\,, & \mathrm{for}\ R \leq \Rb\,. \end{array}
\right.
\end{equation}
Here \Omegab\ is the pattern speed of the bar and \Rb\ is the bar
radius, which is fixed to be 80\,percent of the bar's corotation
radius. The bar is grown smoothly during a time $t_1$, always set to
half of the total integration time in what follows, following the
prescription
\begin{equation}
\Ab(t) = \alpha \, \frac{\vo^2}{3}\,\left(\frac{\Ro}{\Rb}\right)^3
\times \left\{ \begin{array}{lll}
\frac{3}{16}\xi^5 - \frac{5}{8} \xi^3 + \frac{15}{16} \xi + \frac{1}{2}\,, & \xi \equiv 2\frac{t}{t_1}-1\,, & t \leq t_1\,,\\
1\,, & & t \geq t_1\,.\end{array} \right. 
\end{equation}
A comparison of the local velocity distribution obtained by using this
simple bar potential with that resulting from the more realistic bar
potential of \citet{gardner10a} shows that this simple model is
sufficient to model the bar's influence on stellar orbits in the outer
disk.

@@

\citet{dehnen00a} only simulated the velocity distribution at the
position of the Sun for various combinations of the model parameters
defined in the previous Section. Nevertheless, his results on how the
velocity distribution at the Sun changes with the angle between the
Sun-Galactic center line and the bar could also be interpreted as
showing the velocity distribution on the Solar circle for different
Galactocentric azimuths, fixing the bar angle. In this way, these
simulations become predictions for what one might find when one
studies the velocity distribution in regions beyond those probed by
the \hipparcos\ mission.

In \figurename~\ref{fig:rphi2d}, I present the predictions for the
velocity distributions near the Solar circle for the full
360$^{\circ}$ range in Galactocentric azimuth for the fiducial model
from \sectionname~\ref{sec:method}. Only the range $-90^{\circ} \leq
\phi \leq 90^{\circ}$ is shown here, since the barred Galaxy is
symmetric with respect to rotations through 180$^{\circ}$, such that
the predictions for the other side of the Galaxy can be obtained from
those shown by shifting the azimuths by 180$^{\circ}$. The $R = \Ro$
locus is not shown here, but it can be inferred from \figurename~2 in
\citet{dehnen00a}.

It is clear from this \figurename\ that we can trace the Hercules
stream going around the Galaxy and when looking further in toward the
Galactic center or further out. The effect of the bar is also not
confined to a narrow range in radii, rather the effects of the bar and
especially its outer Lindblad resonance are felt through a large range
of radii (0.6 $\leq R/\Ro \leq 1.4$). These effects are strong and
robust predictions of the bar-origin model for the Hercules stream,
especially in regions beyond the Solar radius, where the Hercules
feature can be found significantly removed from the bulk of the
velocity distribution, which might be severely shaped at all locations
by the processes that create the pattern of low-velocity moving groups
in the local velocity distribution (the main mode in
\figurename~\ref{fig:obs}). As I discuss below, \Gaia\ will be able to
verify these predictions.

@@

Rather than wait for a comprehensive survey of the Galactic disk that
measures the full phase-space position of a large number of stars by
measuring proper motions, we can ask whether a targeted survey of
line-of-sight velocities in selected regions of the Galaxy could see a
clear signature of the Hercules feature. That this should be possible
follows from the following argument. The Hercules feature stands out
in the velocity distribution most in its tangential velocity
$\vphi$. On the Solar circle near the Sun we see almost entirely this
Galactocentric tangential velocity in the line-of-sight, so we would
expect to see at least a hint of the Hercules stream in the line-of
sight.

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the tangential velocity. This
\figurename\ is no longer symmetric with respect to shifts of
180$^{\circ}$ degree in azimuth, since the line-of-sight direction
breaks this symmetry. The gray curve is the line-of-sight distribution
one would observe if the targeted stars were in a steady state in an
axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be
great enough that we can test predicted differences in the absolute
number of stars at a given location between the axisymmetric and the
bar-resonance predictions. These differences are also largely a
function of the particular form and parameters of the distribution
function that we choose to represent the stellar disk before bar
formation.

The prospects to detect the Hercules feature are quite good on our
side of the Galaxy. The Hercules feature is especially strong in the
line-of-sight velocity distribution for Galactic longitudes $l \geq
250^{\circ}$. This is because of the fortunate coincidence that for
those lines-of-sight we see velocities in the direction along which
the Hercules streams separates most from the bulk of the velocity
distribution (see \figurename~\ref{fig:rphi2d}) and that at these
azimuths the Hercules stream is at the furthest distance from the rest
of the velocity distribution. We see hardly any effect of the bar on
the line-of-sight velocity distributions on the other side of the
Galaxy ($90^{\circ} \leq \phi \leq 270^{\circ}$). This is due to the
fact that on that side we mostly see the Galactocentric radial
velocity, in which the Hercules feature does not separate clearly from
the bulk of the stars. Even though upcoming surveys such as
\apogee\ have the ability to probe the other side of the Galaxy, for
the purposes of detecting the Hercules stream this is uninteresting.

@@

Finally, in \figurename~\ref{fig:1dvar} I show the influence of
various nuisance and dynamical parameters on the predicted
line-of-sight velocity distribution for the most promising
line-of-sight and distance from the Sun that is more than 3 kpc
away. We see that the effect of distance uncertainties is
negligible. The exact shape and parameters of the stellar distribution
function also only minimally affect the predicted line-of-sight
velocity distribution. Changing the scale-length within its current
uncertainty changes the prediction only in the amplitude of the
effect: The Hercules stream is more prominent for shorter
scale-lengths since then there are relatively more stars in the inner
Galaxy that are brought to the Solar circle by the interaction with
the bar. Using a Shu distribution function \citep{shu69a} instead of
the Dehnen distribution function also gives very similar
predictions. Changing the axisymmetric potential by adjusting the
shape of the rotation curve shifts the peak as expected
\citep[see][]{dehnen00a}, but the Hercules feature nevertheless
remains detectable.

@@

The astrometric \Gaia\ mission will measure the parallaxes and proper
motions of up to one billion stars, most of which will be disk
stars. With parallax and proper motion accuracies down to 10
microarcseconds, \Gaia\ will be able to probe the kinematics of the
disk out to several kpc in all directions
\citep{bailerjones08a}. Crucially, it will provide large samples of
stars in many of the promising regions in
\figurename~\ref{fig:rphi2d}, which will be able to verify the
predictions laid out in this paper. This will allow the dynamical
properties of the bar to be tightly constrained, something that will
be hard for \Gaia\ to do using direct observations of the bulge due to
the large extinction toward the Galactic center \citep{robin05a}.

@@

Since the extinction in the optical bandpasses is rather large in the
directions identified as promising in \figurename~\ref{fig:detect}
\citep[$A_V \approx 8$ mag toward $l \approx
  270^{\circ}$;][]{marshall06a}, spectroscopic surveys in the
near-infrared such as \apogee\ are the best option for seeing the
Hercules feature in the line-of-sight direction. The range of Galactic
longitudes that \apogee\ can target along the disk is $-5^{\circ} \leq
l \leq 250^{\circ}$ (SDSS-III Project Description\footnote{Available
  at \url{http://sdss3.org/collaboration/description.pdf}~.}). From
\figurename~\ref{fig:detect} this limit appears right on the edge of
where one might hope to detect the Hercules feature. Since \apogee\ is
expected to start taking data in 2011 and to finish operations before
the first \Gaia\ data release, \apogee\ could trace the Hercules
stream around the Galaxy before \Gaia\ does. Since \apogee\ will also
obtain detailed chemical abundances for its target stars, this would
also allow for a more detailed analysis of the abundances of stars in
the Hercules stream \citep[following, \eg,][]{Bensby07a,Bovy10a},
which could shed more light on radial mixing in the Galactic disk.

\acknowledgements It is a pleasure to thank Mike Blanton and David
Hogg for helpful discussions and assistance.  Financial support for
this project was provided by the National Aeronautics and Space
Administration (grant NNX08AJ48G) and the National Science Foundation
(grant AST-0908357).

@@

\clearpage
\begin{figure}
\includegraphics[width=\textwidth]{rphi2d.ps}
\caption{Two-dimensional velocity distributions along the Solar
  circle. The axes of each subpanel are the same as in
  \figurename~\ref{fig:obs}. The velocity distributions on the other
  side of the Galaxy are obtained by shifting the azimuths by
  180$^{\circ}$.}\label{fig:rphi2d}
\end{figure}

>>>>

%TO DO:
% - l-range of good directions
% - radial and tangential confusion
% - Hercules in Fig 2.
\documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath}
%\usepackage{color,hyperref}
%\definecolor{linkcolor}{rgb}{0,0,0.25}
%\hypersetup{
%  colorlinks=true,        % false: boxed links; true: colored links
%  linkcolor=linkcolor,    % color of internal links
%  citecolor=linkcolor,    % color of links to bibliography
%  filecolor=linkcolor,    % color of file links
%  urlcolor=linkcolor      % color of external links
%}
\setlength{\emergencystretch}{2em}%No overflowing references
\newcommand{\ie}{i.e.}
\newcommand{\etal}{et al.}
\newcommand{\eg}{e.g.}
\newcommand{\eqnname}{equation}
\newcommand{\figurenames}{\figurename s}
\newcommand{\sectionname}{$\mathsection$}
\newcommand{\tvector}[1]{\boldsymbol{\vec{#1}}}
\newcommand{\vx}{\tvector{x}}
\newcommand{\vv}{\tvector{v}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\Gaia}{\emph{Gaia}}
\newcommand{\apogee}{APOGEE}
\newcommand{\hipparcos}{\emph{Hipparcos}}
\newcommand{\vR}{\ensuremath{v_R}}
\newcommand{\vphi}{\ensuremath{v_{\phi}}}
\newcommand{\vo}{\ensuremath{v_0}}
\newcommand{\Ro}{\ensuremath{R_0}}
\newcommand{\ro}{\Ro}
\newcommand{\Ab}{\ensuremath{A_b}}
\newcommand{\Rb}{\ensuremath{R_b}}
\newcommand{\Omegab}{\ensuremath{\Omega_b}}
\newcommand{\fdehnen}{\ensuremath{f_{\text{Dehnen}}}}
\newcommand{\sigmaR}{\ensuremath{\sigma_R}}
\newcommand{\rE}{\ensuremath{R_e}}
\newcommand{\Lc}{\ensuremath{L_c}}
\newcommand{\Rs}{\ensuremath{R_s}}
\newcommand{\Rsigma}{\ensuremath{R_{\sigma}}}
\newcommand{\Rolr}{\ensuremath{R_{\text{OLR}}}}

@@

\begin{abstract}
It has been proposed that the Hercules stream, a group of co-moving
stars in the Solar neighborhood offset from the bulk of the velocity
distribution, is the result of resonant interactions between stars in
the outer disk and the Galactic bar. So far it has only been seen in
the immediate Solar neighborhood, but any resonance model make a
prediction over a large fraction of the Galactic disk. I predict the
distribution of stellar velocities and changing Hercules signature as
a function of location in the Galactic disk in a simple model for the
Galaxy and the bar that produces the observed Hercules stream. The
Hercules feature is expected to be strong enough to be unambiguously
detected in the distribution of line-of-sight velocities in selected
directions. I identify quantitatively the most promising lines of
sight for detection in line-of-sight velocities using the
Kullback-Leibler divergence between the predictions from the resonance
model and an axisymmetric model; these directions are all at $l
\gtrsim 250^{\circ}$. The predictions presented here are only weakly
affected by distance uncertainties, assumptions about the distribution
function in the stellar disk, and the details of the Galactic
potential including the effect of spiral structure. \Gaia\ and future
spectroscopic surveys of the Galactic disk will be able to robustly
test the origin of the Hercules stream and constrain the properties of
the Galactic bar.
\end{abstract}

@@

The velocity distribution of nearby stars contains various large
overdensities of co-moving stars that cannot be explained as the
vestiges of clumpy star formation \citep[\eg,][]{Bovy10a}. In
particular, the Hercules moving group, a group of stars significantly
offset from the bulk of the observed velocity distribution
\citep[\eg,][see \figurename~\ref{fig:obs}]{Dehnen98b,Bovy09a},
displays a wide range of ages and metallicities
\citep{Blaauw70a,raboud98a,caloi99a,Bensby07a,Bovy10a}. A compelling
model has the Hercules stream originating through resonant
interactions of stars in the outer disk with the bar in the central
region of the Galaxy \citep{dehnen00a,fux01a}, but there is no
confirmation of this picture beyond the locally observed stellar
kinematics that it was proposed to explain. An investigation of the
Hercules members' color--magnitude diagram found no evidence for any
significant metallicity anomaly with respect to other local stars
\citep{Bovy10a}, even though such a difference should be expected as
the Hercules stars originate from a few kiloparsec toward the Galactic
center where the average metallicity is higher than in the Solar
neighborhood. The absence of the predicted metallicity anomaly could
potentially be explained, \eg, through radial mixing
\citep{sellwood02a}, but it is clear that chemodynamical modeling of
the Galactic disk is not currently up to the task of separating out
these effects.

In this Letter I propose that a robust and unambiguous test of the
bar-resonance model for the Hercules stream is a search for the
distinct pattern it predicts as we trace the velocity distribution of
stars near the Solar radius around the Galaxy. \Gaia\ is in a good
position to have the final word on this, but I also show that the
Hercules feature can be detected in the line-of-sight velocity
distribution in selected directions on the sky. Clear signatures are
predicted to be detectable in regions a few kiloparsec from the Sun,
allowing for a clean verification of the bar-resonance model for the
Hercules stream.

@@

To simulate the effect of the bar on stellar orbits near the Solar
circle, we follow the approach of \citet{dehnen00a}. For each fixed
position in the Galaxy, we evaluate the velocity distribution function
at values for the radial---toward the Galactic center---and
tangential---in the direction of Galactic rotation---velocity by
backward-integrating these velocities in the Galactic potential to
obtain the intial orbit before bar formation. We posit that at this
epoch the old stellar disk was in a steady-state that can be described
by a simple distribution function that is a function of the energy and
angular momentum of the orbit alone. It follows from the collisionless
Boltzmann equation that the value of the velocity distribution at the
present position and velocity is equal to that of the steady-state
distribution function before bar formation evaluated at the initial
orbit.

Since the Hercules stream is only apparent in the planar motions in
the disk \citep[\eg,][]{Bovy09a}, we only consider a two-dimensional
model for the Galaxy. We use a simple power-law rotation curve to
model the axisymmetric Galactic potential
\begin{equation}
v_c(R) = \vo(R/\Ro)^\beta \,,
\end{equation}
where \Ro\ is the distance from the Sun to the Galactic center. For
most of the simulations below we specify this model further to a flat
rotation curve ($\beta = 0$). The model for the bar is given by the
following potential
\begin{equation}
\Phi_b(R,\phi) = \Ab(t) \cos [2(\phi - \Omegab t)] \times
\left\{ \begin{array}{ll} -(\Rb/R)^3\,, & \mathrm{for}\ R \geq
  \Rb\,,\\ (R/\Rb)^3-2\,, & \mathrm{for}\ R \leq \Rb\,. \end{array}
\right.
\end{equation}
Here \Omegab\ is the pattern speed of the bar and \Rb\ is the bar
radius, which is fixed to be 80\,percent of the bar's corotation
radius. The bar is grown smoothly during a time $t_1$, always set to
half of the total integration time in what follows, using the
prescription
\begin{equation}
\Ab(t) = \alpha \, \frac{\vo^2}{3}\,\left(\frac{\Ro}{\Rb}\right)^3
\times \left\{ \begin{array}{lll}
\frac{3}{16}\xi^5 - \frac{5}{8} \xi^3 + \frac{15}{16} \xi + \frac{1}{2}\,, & \xi \equiv 2\frac{t}{t_1}-1\,, & t \leq t_1\,,\\
1\,, & & t \geq t_1\,.\end{array} \right. 
\end{equation}
A comparison of the local velocity distribution obtained by using this
simple bar potential with that resulting from the more realistic one
of \citet{gardner10a} shows that this simple potential is sufficient
to model the bar's influence on stellar orbits in the outer disk.

@@

\citet{dehnen00a} only simulated the velocity distribution at the
position of the Sun for various combinations of the model parameters
defined in the previous Section. Nevertheless, his results on how the
velocity distribution at the Sun changes with the angle between the
Sun-Galactic center line and the bar could also be interpreted as
showing the velocity distribution on the Solar circle for different
Galactocentric azimuths, fixing the bar angle. In this way, these
simulations can make predictions for what we might find when we study
the velocity distribution in regions beyond those probed by the
\hipparcos\ mission.

In \figurename~\ref{fig:rphi2d}, I present the predictions for the
two-dimensional velocity distributions near the Solar circle for the
full 360$^{\circ}$ range in Galactocentric azimuth for the fiducial
model from \sectionname~\ref{sec:method}. Only the range $-90^{\circ}
\leq \phi \leq 90^{\circ}$ is shown here, since the barred Galaxy is
symmetric with respect to rotations through 180$^{\circ}$; the
predictions for the other side of the Galaxy can be obtained from
those shown by shifting the azimuths by 180$^{\circ}$. The predicted
velocity distributions on the Solar circle are not shown here, but
they can be inferred from \figurename~2 in \citet{dehnen00a}.

It is clear from this \figurename\ that we can trace the Hercules
stream going around the Galaxy and when looking further in toward the
Galactic center or further out. Furthermore, the effect of the bar is
not confined to a narrow range in radii, but rather the effects of the
bar and especially its outer Lindblad resonance are felt through a
large range of radii (0.6 $\leq R/\Ro \leq 1.4$). These effects are
strong and robust predictions of the bar-origin model for the Hercules
stream. This is especially the case in regions beyond the Solar
radius. There the Hercules feature can be found significantly removed
from the bulk of the velocity distribution, which might be severely
shaped at all locations by the processes that create the pattern of
low-velocity moving groups in the local velocity distribution (the
main mode in \figurename~\ref{fig:obs}). As I discuss below, \Gaia\
will be able to verify these predictions.

@@

Rather than wait for a comprehensive survey of the Galactic disk that
measures the full phase-space position of a large number of stars by
measuring proper motions, we can ask whether a targeted survey of
line-of-sight velocities in selected regions of the Galaxy could
detect a clear signature of the Hercules feature? The Hercules feature
stands out in the velocity distribution most in its tangential
velocity $\vphi$. On the Solar circle near the Sun the line-of-sight
velocity is almost entirely this Galactocentric tangential velocity,
so we would expect to see at least a hint of the Hercules stream in
the line-of sight.

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the tangential velocity. This \figurename\
is no longer symmetric with respect to shifts of 180$^{\circ}$ degree
in azimuth, since the line-of-sight direction breaks this
symmetry. The gray curve is the line-of-sight distribution we would
observe if the targeted stars were in a steady state in an
axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be
great enough that we can test predicted differences in the absolute
number of stars at a given location between the axisymmetric and the
bar-resonance predictions. These differences are also largely a
function of the particular form and parameters of the distribution
function that we choose to represent the stellar disk before bar
formation.

The prospects to detect the Hercules feature are quite good on our
side of the Galaxy. The Hercules feature is especially strong in the
line-of-sight velocity distribution for Galactic longitudes $l \geq
250^{\circ}$. This is because of the fortunate coincidence that for
those lines-of-sight we see velocities in the direction along which
the Hercules streams separates most from the bulk of the velocity
distribution (see \figurename~\ref{fig:rphi2d}) and that at these
azimuths the Hercules stream is at the furthest distance from the rest
of the velocity distribution. We see hardly any effect of the bar on
the line-of-sight velocity distributions on the other side of the
Galaxy ($90^{\circ} \leq \phi \leq 270^{\circ}$). On that side we
mostly see the Galactocentric radial velocity in the line-of-sight, in
which the Hercules feature does not separate clearly from the bulk of
the stars. Even though upcoming surveys such as \apogee\ have the
ability to probe the other side of the Galaxy, for the purposes of
detecting the Hercules stream this is uninteresting.

@@

Finally, in \figurename~\ref{fig:1dvar} I show the influence of
various nuisance and dynamical parameters on the predicted
line-of-sight velocity distribution for the most promising region that
is more than 3 kpc away. We see in this \figurename\ that the effect
of distance uncertainties is negligible. The exact shape and
parameters of the stellar distribution function also only minimally
affect the predicted line-of-sight velocity distribution. Changing the
scale-length within its current uncertainty changes the prediction
only in the amplitude of the effect: The Hercules stream is more
prominent for shorter scale-lengths since then there are relatively
more stars in the inner Galaxy that are brought to the Solar circle by
the interaction with the bar. Using a Shu distribution function
\citep{shu69a} instead of the Dehnen distribution function also gives
very similar predictions. Changing the axisymmetric potential by
adjusting the shape of the rotation curve shifts the peak as expected
\citep[see][]{dehnen00a}, but the Hercules feature nevertheless
remains detectable.

@@

The astrometric \Gaia\ mission will measure the parallaxes and proper
motions of up to one billion stars, most of which will be disk
stars. With parallax and proper motion accuracies down to 10
microarcseconds, \Gaia\ will be able to probe the kinematics of the
disk out to several kpc in all directions
\citep{bailerjones08a}. Crucially, it will provide large samples of
stars in many of the promising regions in
\figurename~\ref{fig:rphi2d}, which will be able to verify the
predictions laid out in this paper. This will also allow the dynamical
properties of the bar to be tightly constrained, something that will
be hard for \Gaia\ to do using direct observations of the bulge due to
the large extinction toward the Galactic center \citep{robin05a}.

@@

Since the extinction in the optical bandpasses is rather large in the
directions identified as promising in \figurename~\ref{fig:detect}
\citep[$A_V \approx 8$ mag toward $l \approx
270^{\circ}$;][]{marshall06a}, spectroscopic surveys in the
near-infrared such as \apogee\ are the best option for seeing the
Hercules feature in the line-of-sight direction. The range of Galactic
longitudes that \apogee\ can target along the disk is $-5^{\circ} \leq
l \leq 250^{\circ}$ (SDSS-III Project Description\footnote{Available
at \url{http://sdss3.org/collaboration/description.pdf}~.}). From
\figurename~\ref{fig:detect} this limit appears right on the edge of
where we might hope to detect the Hercules feature. Since \apogee\ is
expected to start taking data in 2011 and to finish operations before
the first \Gaia\ data release, \apogee\ could trace the Hercules
stream around the Galaxy before \Gaia\ does. Since \apogee\ will also
obtain detailed chemical abundances for its target stars, this would
also allow for a more detailed analysis of the abundances of stars in
the Hercules stream \citep[following, \eg,][]{Bensby07a,Bovy10a},
which could shed more light on radial mixing in the Galactic disk.

\acknowledgements It is a pleasure to thank Mike Blanton and David
W.~Hogg for helpful discussions and assistance.  Financial support for
this project was provided by the National Aeronautics and Space
Administration (grant NNX08AJ48G) and the National Science Foundation
(grant AST-0908357).

@@

\clearpage
\begin{figure}
\includegraphics[width=\textwidth]{rphi2d.ps}
\caption{Two-dimensional velocity distributions along the Solar
  circle. The $x$-axis in each subplot is the Galactocentric radial
  velocity and the $y$-axis is the Galactocentric tangential velocity;
  the axis-ranges of each subplot are the same as in
  \figurename~\ref{fig:obs}. The velocity distributions on the other
  side of the Galaxy are obtained by shifting the azimuths by
  180$^{\circ}$.}\label{fig:rphi2d}
\end{figure}
